Revealing Folk Schemas of Musical Genre and Social Category Associations via Relational and Geometric Methods\thanks{Reserved for acknowledgments}

### 2025-06-21 manuscript.tex

\subsection*{Averaged Genre Associations}
Figure~\ref{fig:main} shows a heatmap plot of the proportion of respondents associating each genre with each of the fifteen social labels included in the survey, reproducing one of the key results from \citet[Table 3]{lizardo2016cultural-aaa}. As we can see, respondents use social labels associated with gender fairly indiscriminately, associating it with each genre at a fairly high rate. Racial categories are used in a more discriminating manner, reproducing (on the average) standard stereotypes of genre audiences. Rock and traditional ``highbrow'' genre categories are associated with white audiences, Rap, Blues, R\&B, Reggae, and Gospel with Black audiences, and Latin music with Hispanic audiences. The category of Asian is not strongly associated with any genre. Both age and class/education categories also reproduce expected stereotypes. For instance, ,Rap and Metal are considered ``young'' genres, while Classic Rock and Country are thought of as ```middle aged.'' In the same way, Classical and Opera audiences are considered educated and Upper Class, while Bluegrass and Country audiences are considered not college educated and working or lower class).

@@

\subsection*{Cognitive Two Mode Networks}
The data used to construct Figure ~\ref{fig:main} consists of averages of more than two thousand binary matrices---one for each respondent---of the same dimensions as~\ref{fig:main} provided by each respondent. As such, the averaging hides as much as it reveals, even as it produced intuitive results. The methodological approach taken here treats this rich disaggregated association data as cognitive two-mode data, conceptually analogous to Krackhardt-style data used in one-mode networks \citep{batchelder1997consensus-f5c, kumbasar1994systematic-213}, where each person provides their perceived associations between musical genres and social labels. In terms of recent work in measuring culture, each individual two-mode data table records the (explicit) schematic structure of the way individuals perceive each genre category relative to the set of social labels \citep{hunzaker2019mapping-280}.

@@

>>>>

\subsection*{Averaged Genre Associations}
Figure~\ref{fig:main} shows a heatmap plot of the proportion of respondents associating each genre with each of the fifteen social labels included in the survey, reproducing one of the key results from \citet[Table 3]{lizardo2016cultural-aaa}. As we can see, respondents use social labels associated with gender fairly indiscriminately, associating them with each genre at a fairly high rate. Racial categories are used in a more discriminating manner, reproducing (on average) standard stereotypes of genre audiences. Rock and traditional ``highbrow" genre categories are associated with white audiences, while Rap, Blues, R\&B, Reggae, and Gospel are associated with Black audiences, and Latin music is associated with Hispanic audiences. The category of Asian is not strongly associated with any genre. Both age and class/education categories also reproduce expected stereotypes. For instance, Rap and Metal are considered ``young'' genres. At the same time, Classic Rock and Country are thought of as ```middle-aged.'' In the same way, Classical and Opera audiences are considered educated and Upper Class, while Bluegrass and Country audiences are considered not college educated and working or lower class.

@@

\subsection*{Cognitive Two Mode Networks}
The data used to construct Figure ~\ref{fig:main} consists of averages of more than two thousand binary matrices---one for each respondent---of the same dimensions as~\ref{fig:main} provided by each respondent. As such, the averaging hides as much as it reveals, even as it produces intuitive results. The methodological approach taken here treats this rich disaggregated association data as cognitive two-mode data, conceptually analogous to Krackhardt-style data used in one-mode networks \citep{batchelder1997consensus-f5c, kumbasar1994systematic-213}, where each person provides their perceived associations between musical genres and social labels. In terms of recent work on measuring culture, each person-level two-mode data table records the (explicit) schematic structure of how individuals perceive each genre category relative to the set of social labels \citep{hunzaker2019mapping-280}.

@@

The SDSM works as follows: In the first step, a generalized linear model for binary outcomes is used to predict the probability of a tie between each genre and label in the two-mode network, using the number of times a genre was associated with social label and the number of times a social label was associated with a genre (the degrees of the nodes in each mode) and their interaction as predictors. From the predicted probabilities derived from this model (where the unit of analysis is the genre/label pair), we create an ensemble of two-mode networks, generate genre-to-genre and label-to-label projections for each, and compute the percentage of times that the projected edge weight between pairs of genres and pairs of labels is smaller than that observed in the original network. We then create a binary link between two genres and two labels (indicating perceived similarity) if this percentage falls below a certain threshold. In these data, the usual $p < 0.05$ criterion proves to be too strong (generating mostly empty backbones) for each genre-to-genre and label-to-label projection. Accordingly, I use a less stringent criterion of $p < 0.45$.
